---
title: "Hyperparameter tuning of the Random Forest Classifier"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>", dpi = 500)

suppressPackageStartupMessages(library(tidymodels))
suppressPackageStartupMessages(library(depmixS4))

suppressPackageStartupMessages(library(kableExtra))

source(here::here("start.R"))
theme_set(theme_minimal())
set.seed(0)
```

The purpose of this notebook is the comprehensively tune the hyperparameters of a Random Forest (RF) Classifier for detecting states of a push-up.
The [*Hidden Markov Model-based learning pipelines*](05_008_hmm_pipelines.md) notebook demonstrated that a RF is likely the best model for this task, though did not pursue full optimization of the classifier.


## Data

The data used here was cached in [*Hidden Markov Model-based learning pipelines*](05_008_hmm_pipelines.md).

```{r}
processed_pushup_data <- readRDS(
  here(file.path("cache", "hmm_processed_pushup_data.rds"))
)
processed_pushup_data
```

## Hyperparameter tuning

**To-Do**: learn about the tuning process available in TidyModels
