---
title: "Hidden Markov Modeling"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>", dpi = 500)

suppressPackageStartupMessages(library(depmixS4))

source(here::here("start.R"))
theme_set(theme_minimal())
set.seed(0)
```


# Exploratoring the data

Load in some push-up data and explore the numbers.

```{r}
all_data_files <- get_data_file_names(data_dir)
raw_pushup_data <- read_watch_data(all_data_files[2])

raw_pushup_telemetry_data <- raw_pushup_data$telemetry_data
```

Plot the data over time.

```{r}
line_plot_telemetry <- function(df, x = value) {
  df %>%
    ggplot(aes(idx, {{ x }})) +
    facet_wrap(~ motion, ncol = 1, scales = "free_y") +
    geom_line(aes(color = axis), alpha = 0.7) +
    scale_color_brewer(type = "qual", palette = "Dark2")
}

line_plot_telemetry(raw_pushup_telemetry_data)
```

Subset just the portions of the data where the push-ups are.
This is some usefuly "clean" data for initial model fitting and testing.

```{r}
clean_pushup_data <- raw_pushup_telemetry_data %>%
  filter(between(idx, 250, 1400))
line_plot_telemetry(clean_pushup_data) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold")
  ) +
  labs(title = "Subsetted push-up data")
```

## Data transformations

### Smoothened data

The `ksmooth` function from the 'stats' package has 2 kernel methods: *normal* and *box*.

```{r}
clean_pushup_data %.% {
  group_by(axis, motion)
  mutate(smooth_value = ksmooth(date, value, kernel = "normal")$y)
  ungroup()
} %>%
  line_plot_telemetry(smooth_value)
```

```{r}
clean_pushup_data %.% {
  group_by(axis, motion)
  mutate(smooth_value = ksmooth(date, value, kernel = "box")$y)
  ungroup()
} %>%
  line_plot_telemetry(smooth_value)
```

In general, smoothing the data seems to have a larger effect on the acceleration data.
Because the acceleration seems to oscillate back and forth very rapidly, the smoothing just crishes it back towards 0.

```{r}
clean_pushup_data %.% {
  group_by(axis, motion)
  mutate(smooth_value = ksmooth(date, value, kernel = "normal")$y)
  ungroup()
  filter(between(idx, 600, 1000))
  pivot_longer(
    -c(date, idx, axis, motion),
    names_to = "transformation",
    values_to = "value"
  )
  filter(motion == "acceleration")
} %>%
  ggplot(aes(idx, value)) +
  facet_wrap(~ axis, ncol = 1) +
  geom_line(aes(color = axis, alpha = transformation, size = transformation)) +
  scale_color_brewer(type = "qual", palette = "Set1") +
  scale_size_manual(values = c(2, 1)) +
  scale_alpha_manual(values = c(0.7, 0.4))
```

Perhaps, a running measure of values could be informative.

```{r}
# Performs an action `fxn` on a running basis over `x`.
running_fxn <- function(x, fxn, n = 5) {
  y <- x
  for (i in seq(1, length(x))) {
    x_i <- c()
    for (j in seq(-n, n)) {
      idx <- i + j
      if (idx < 1 | idx > length(x)) { next }
      x_i <- c(x_i, x[[idx]])
    }
    y[[i]] <- fxn(x_i)
  }
  return(y)
}
```

Below are plots of the acceleration data with a few different transformations.
A possible conclusion is that it may not be the actual acceleration values that matter, but how variable the data is.
The `smooth_abs_max` value looks very promising. 
It is the smoothened version of `running_abs_max` which is maximum absolute value on a running-basis (± 10 positions on either side).

```{r}
test_smoothing_trans <- function(df, x = value) {
  df %.% {
    group_by(axis, motion)
    mutate(
      running_var = running_fxn({{ x }}, fxn = var, n = 10),
      running_abs_max = running_fxn(abs({{ x }}), fxn = max, n = 10),
      smooth_abs_value = ksmooth(date, abs({{ x }}), kernel = "box")$y,
      smooth_abs_max = ksmooth(date, running_abs_max, kernel = "box")$y
    )
  }
}


clean_pushup_data %.% {
  test_smoothing_trans(value)
  filter(between(idx, 600, 800))
  pivot_longer(
    -c(date, idx, axis, motion),
    names_to = "transformation",
    values_to = "value"
  )
  filter(motion == "acceleration")
}%>%
  ggplot(aes(idx, value)) +
  facet_grid(axis ~ transformation) +
  geom_line(aes(color = transformation), alpha = 0.8, size= 1.2) +
  scale_color_brewer(type = "qual", palette = "Set2", guide = FALSE) 
```

The smoothing method on the cleaned push-up data.

```{r}
clean_pushup_data %>%
  test_smoothing_trans(value) %>%
  line_plot_telemetry(smooth_abs_max)
```

The smoothing method on the raw push-up data (all of the data, not just the clipped middle portion).

```{r}
raw_pushup_telemetry_data %>% 
  test_smoothing_trans(value) %>%
  line_plot_telemetry(smooth_abs_max)
```

```{r}
apply_smoothing_trans <- function(df, x = value, rolling_n = 10) {
  df %.% {
    group_by(axis, motion)
    mutate(
      running_abs_max = running_fxn(abs({{ x }}), fxn = max, n = rolling_n),
      smooth_abs_max = ksmooth(date, running_abs_max, kernel = "box")$y
    )
  }
}
```

Trying different rolling-window sizes.

```{r}
raw_pushup_rolling_ns <- raw_pushup_telemetry_data %>% 
  apply_smoothing_trans(value, rolling_n = 5) %>%
  select(date, idx, axis, motion, value, smooth_abs_max5 = smooth_abs_max) %>%
  apply_smoothing_trans(value, rolling_n = 10) %>%
  select(-running_abs_max) %>%
  rename(smooth_abs_max10 = smooth_abs_max) %>% 
  apply_smoothing_trans(value, rolling_n = 20) %>%
  select(-running_abs_max) %>%
  rename(smooth_abs_max20 = smooth_abs_max) %>% 
  apply_smoothing_trans(value, rolling_n = 40) %>%
  select(-running_abs_max) %>%
  rename(smooth_abs_max40 = smooth_abs_max) %>% 
  apply_smoothing_trans(value, rolling_n = 60) %>%
  select(-running_abs_max) %>%
  rename(smooth_abs_max60 = smooth_abs_max) %>% 
  apply_smoothing_trans(value, rolling_n = 80) %>%
  select(-running_abs_max) %>%
  rename(smooth_abs_max80 = smooth_abs_max) %>% 
  apply_smoothing_trans(value, rolling_n = 100) %>%
  select(-running_abs_max) %>%
  rename(smooth_abs_max100 = smooth_abs_max)

raw_pushup_rolling_ns %.% {
  rename(value0 = value)
  pivot_longer(-c(date, idx, axis, motion))
  mutate(rolling_n = as.numeric(str_extract(name, "[:digit:]+")))
  filter(motion == "acceleration")
} %>%
  ggplot(aes(date, value)) +
  facet_wrap(~ axis, ncol = 1, scales = "free") +
  geom_line(aes(color = factor(rolling_n)), size = 0.6, alpha = 0.7) +
  theme(legend.position = "right") +
  labs(x = "time point",
       y = "value",
       color = "rolling-window\nsize")
```
A rolling-window of size 10-20 should be good.
I'll leave the default value at 10.

### Scaled data

```{r}
zscale <- function(x, na.rm = TRUE) {
  (x - mean(x, na.rm = na.rm)) / sd(x, na.rm = na.rm)
}


clean_pushup_data %.% {
  group_by(axis, motion)
  mutate(
    scaled_value = zscale(value)
  )
  ungroup()
} %>%
  line_plot_telemetry(scaled_value)
```

Applying the smoothing method from above to scaled data.

```{r}
clean_pushup_data %.% {
  group_by(axis, motion)
  mutate(
    scaled_value = zscale(value)
  )
  ungroup()
  apply_smoothing_trans(scaled_value)
} %>%
  line_plot_telemetry(smooth_abs_max)
```

The scaled smooth data looks very promising.

```{r}
raw_pushup_telemetry_data %.% {
  group_by(axis, motion)
  mutate(
    scaled_value = zscale(value)
  )
  ungroup()
  apply_smoothing_trans(scaled_value)
} %>%
  line_plot_telemetry(smooth_abs_max)
```

```{r}
apply_scale_trans <- function(df, x = value) {
  df %.% {
    group_by(axis, motion)
    mutate(scaled_value = zscale({{ x }}))
    ungroup()
  }
}
```

### Log-transformation

The log transform could help adjust the data to appear more normal.
From the plots below, the scaled data is pretty normal, though the smoothing process disrupts that.
Taking the logarithm seems to help, though it may come down to experimenting with the model fitting process.

```{r}
clean_pushup_data %.% {
  apply_scale_trans(value)
  apply_smoothing_trans(scaled_value)
  log_smooth_value = log(smooth_abs_max + 0.001)
  
  select(date, idx, axis, value, scaled_value, smooth_abs_max, log_smooth_value)
  pivot_longer(-c(date, idx, axis),
               names_to = "trans_type",
               values_to = "value")
} %>%
  ggplot(aes(value)) +
  facet_wrap(~ trans_type, scales = "free", ncol = 2) +
  geom_density(aes(color = axis, fill = axis), size = 0.9, alpha = 0.3)
```

### A single function for applying all data transforms

```{r}
apply_data_transforms <- function(df, x = value, rolling_n = 10) {
  df %.% {
    apply_scale_trans({{ x }})
    apply_smoothing_trans(scaled_value, rolling_n = rolling_n)
    log_smooth_value = log(smooth_abs_max + 0.0001)
  }
}
```


## Exploration of Hidden Markov Models (HMM)

Fit 3 different types of models: one with all of the data, one with just the acceleration data, and one with just the attitude data.

```{r}
pivot_telemetry_data <- function(telemetry_data, x = value) {
  telemetry_data %>%
    pivot_wider(
      c(date, idx),
      names_from = axis,
      values_from = {{ x }}
    )
}


construct_full_telemetry_hmm_model <- function(d, nstates) {
  depmix(
    list(
      x ~ 1,
      y ~ 1,
      z ~ 1,
      pitch ~ 1,
      roll ~ 1,
      yaw ~ 1
    ),
    nstates = nstates,
    family = list(
      gaussian(), gaussian(), gaussian(),
      gaussian(), gaussian(), gaussian()
    ),
    data = d
  )
}


construct_accel_telemetry_hmm_model <- function(d, nstates) {
  depmix(
    list(
      x ~ 1,
      y ~ 1,
      z ~ 1
    ),
    nstates = nstates,
    family = list(gaussian(), gaussian(), gaussian()),
    data = d
  )
}


construct_attitude_telemetry_hmm_model <- function(d, nstates) {
  depmix(
    list(
      pitch ~ 1,
      roll ~ 1,
      yaw ~ 1
    ),
    nstates = nstates,
    family = list(gaussian(), gaussian(), gaussian()),
    data = d
  )
}
```

```{r}
pushup_data_wide <- pivot_telemetry_data(clean_pushup_data)

hmm_construction_functions <- c(
  "full" = construct_full_telemetry_hmm_model,
  "accel_only" = construct_accel_telemetry_hmm_model,
  "attitude_only" = construct_attitude_telemetry_hmm_model
)


hmm_cleaned_models <- tibble(
  nstates = rep(c(2, 3), 3),
  fxn_name = rep(names(hmm_construction_functions), each= 2),
  fxn = rep(hmm_construction_functions, each = 2)
) %>%
  mutate(
    model = map2(fxn, nstates, ~ .x(d = pushup_data_wide, nstates = .y)),
    fit = map(model, fit),
    aic = map_dbl(fit, AIC),
    bic = map_dbl(fit, BIC)
  )

hmm_cleaned_models %>%
  arrange(aic, bic)
```

```{r}
plot_hmm_results <- function(hmm_fit) {
  posterior(hmm_fit) %>%
    as_tibble() %>%
    mutate(idx = row_number()) %>%
    pivot_longer(-c(idx, state)) %>%
    ggplot(aes(x = idx, y = value, color = name)) +
    facet_grid(name ~ .) +
    geom_line(size = 1.2, alpha = 0.8) +
    scale_color_brewer(type = "qual", palette = "Set1") +
    scale_y_continuous(breaks = c(0, 0.5, 1)) +
    theme(
      plot.title = element_text(hjust = 0.5, size = 11, face = "bold"),
    ) +
    labs(
      x = "data point",
      y = "probability of state",
      color = "state",
      title = "Hidden Markov Model states"
    )
}


data_and_hmm_plot <- function(model_data, hmm_fit, x = value) {
  data_p <- model_data %>%
    mutate(motion_type = str_to_title(axis)) %>%
    line_plot_telemetry(x = {{ x }}) +
    labs(
      x = NULL,
      y = "value"
    ) +
    theme(
      strip.text = element_text(size = 11, face = "bold")
    )

  hmm_p <- plot_hmm_results(hmm_fit)

  data_p / hmm_p
}
```

```{r}
get_best_model <- function(model_df) {
  model_df$fit[which.min(model_df$aic)][[1]]
}

best_cleaned_model <- get_best_model(hmm_cleaned_models)
summary(best_cleaned_model)
```

```{r}
data_and_hmm_plot(clean_pushup_data, best_cleaned_model) +
  plot_layout(heights = c(2, 1))
```


```{r}
raw_pushup_wide <- pivot_telemetry_data(raw_pushup_telemetry_data)

hmm_raw_models <- tibble(
  nstates = c(2:5),
  fxn = rep(c(construct_full_telemetry_hmm_model), 4)
) %>%
  mutate(
    model = map2(fxn, nstates, ~ .x(d = raw_pushup_wide, nstates = .y)),
    fit = map(model, fit),
    aic = map_dbl(fit, AIC),
    bic = map_dbl(fit, BIC)
  )

hmm_raw_models
```

```{r}
best_raw_model <- get_best_model(hmm_raw_models)
summary(best_raw_model)
```
```{r}
data_and_hmm_plot(raw_pushup_telemetry_data, best_raw_model) +
  plot_layout(heights = c(2, 1))
```

### Fitting with transformed data

```{r}
clean_trans_pushups <- clean_pushup_data %>% 
  apply_data_transforms(value)

raw_trans_pushups <- raw_pushup_telemetry_data %>% 
  apply_data_transforms(value)
```

```{r}
clean_trans_pushups %>%
  line_plot_telemetry(log_smooth_value) +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(title = "Transformed cleaned push-up data")
```

```{r}
raw_trans_pushups %>%
  line_plot_telemetry(log_smooth_value) +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(title = "Transformed raw push-up data")
```

```{r}
raw_pushup_wide <- pivot_telemetry_data(raw_trans_pushups, 
                                        x = smooth_abs_max)

hmm_raw_models <- tibble(
  nstates = c(2:5),
  fxn = rep(c(construct_full_telemetry_hmm_model), 4)
) %>%
  mutate(
    model = map2(fxn, nstates, ~ .x(d = raw_pushup_wide, nstates = .y)),
    fit = map(model, fit),
    aic = map_dbl(fit, AIC),
    bic = map_dbl(fit, BIC)
  )

hmm_raw_models %>%
  arrange(aic, bic)
```

```{r}
data_and_hmm_plot(raw_trans_pushups,
                  get_best_model(hmm_raw_models),
                  x = smooth_abs_max) +
  plot_layout(heights = c(2, 1))
```


```{r}
data_and_hmm_plot(raw_trans_pushups,  
                  hmm_raw_models$fit[hmm_raw_models$nstates == 4][[1]],
                  x = smooth_abs_max) +
  plot_layout(heights = c(2, 1))
```

```{r}
clean_pushup_wide <- pivot_telemetry_data(clean_trans_pushups, 
                                          x = smooth_abs_max)

hmm_clean_models <- tibble(
  nstates = c(2:7),
  fxn = rep(c(construct_full_telemetry_hmm_model), length(2:7))
) %>%
  mutate(
    model = map2(fxn, nstates, ~ .x(d = clean_pushup_wide, nstates = .y)),
    fit = map(model, fit),
    aic = map_dbl(fit, AIC),
    bic = map_dbl(fit, BIC)
  )

hmm_clean_models %>%
  arrange(aic, bic)
```

```{r}
data_and_hmm_plot(clean_trans_pushups,  
                  hmm_clean_models$fit[hmm_clean_models$nstates == 2][[1]],
                  x = smooth_abs_max) +
  plot_layout(heights = c(2, 1))
```

The smoothed data seems to work the best.
I think applying the log-transform squishes too many of the distinguishing features out of the data.

---

## Conclusions

1. Smoothing and scaling the data seems to help, though a log-transformation squishes the data more than desired.
2. An HMM can identify the states of a push-up, though it has difficulty with the data from before and after the push-ups begin.
